---
title: "Team Poster Project: Singapore Rainfall Visualisation Redesign"
author: "Team Coral"
format:
  html:
    toc: true
    embed-resources: true
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(lubridate)
library(janitor)
library(viridis)
library(patchwork)
library(scales)
library(ggrepel)
library(zoo)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  dpi = 160
)
```

## 1 Project Overview

This project redesigns a published rainfall visualisation about Singapore's rainfall patterns. The redesign uses station-level daily rainfall observations for Changi from 1980 to 2026.

| Project Component | Redesigned Version |
|------------------------|-----------------------------------------------|
| Original visualisation | "Distribution of daily (left) and monthly (right) rainfall in Singapore; 1983 - 2019" (Chart 5) |
| Redesign topic | Month-by-year flood-stress patterns from heavy and extreme daily rainfall |
| Raw data file | `MSS_Master_Daily_S24_1980_2026.csv` |
| Main variables used | `year`, `month`, `day`, `station`, `daily_rainfall_total_mm` |
| Improved visualisation | Year-month flood-stress matrix using daily heavy-rain and extreme-rain thresholds |

### 1.1 Original Visualisation and Critique

The original Chart 5 in *Visualising Singapore's Changing Weather Patterns: 1983–2019* by Chua Chin Hon was designed to show the distribution of daily (left) and monthly (right) rainfall. It plotted every daily observation and monthly total chronologically as a scatter plot over time. The $y$-axes were inverted, and points were mapped to size and color based on rainfall intensity.

Our critique identifies five main weaknesses:

1.  **Overplotting and Noise:** Daily rainfall is highly skewed (most days are dry). Scatterplotting 13,000+ points results in a dense block of overlapping points at $0\text{ mm}$ that obscures the actual distribution.
2.  **Inverted Vertical Axis:** Larger rainfall amounts are plotted downwards. This is counterintuitive for readers and makes it difficult to read and compare quantities.
3.  **Incomparable Shared Y-Axis:** The subplots share a y-axis. Daily rainfall ($0\text{--}150\text{ mm}$) has a completely different range from monthly rainfall ($0\text{--}700\text{ mm}$), causing the daily plot to be heavily compressed at the top.
4.  **Redundant Channel Mapping:** Mapping rainfall to position, color (hue), and point size simultaneously creates visual clutter without conveying additional variables.
5.  **No Planning Threshold Context:** Plotting raw points chronologically does not show how often rainfall crosses practical heavy-rain or extreme-rain thresholds in each month and year.

### 1.2 Design Brief

Our plan for this redesign is for public climate awareness and urban drainage planning. The goal is to show when heavy-rain and extreme-rain days cluster across months and years, so readers can compare seasonal flood-stress patterns without scanning thousands of daily points.

*Main revisions:*

- **Year-Month Matrix Layout** to preserve both the long-term record and month-by-month seasonality in one view.
- **Heavy-Rain Day Classes** to encode the monthly count of days with at least $50\text{ mm}$ of rainfall.
- **Extreme-Rain Markers** to flag days with at least $100\text{ mm}$ of rainfall without hiding the heavy-rain count.
- **Complete-Year Filtering** to use 1983--2025 and exclude the incomplete 2026 record from year-to-year comparison.

------------------------------------------------------------------------

## 2 Import the Rainfall Data

The raw dataset is stored as a CSV file. The code below searches for the file in the project root first and then in `data/raw/` to ensure the report renders reliably.

```{r}
#| label: import-rainfall-data
find_project_root <- function(start_dir = getwd()) {
  current_dir <- normalizePath(start_dir, winslash = "/", mustWork = TRUE)

  repeat {
    has_source_csv <- file.exists(
      file.path(current_dir, "MSS_Master_Daily_S24_1980_2026.csv")
    )
    has_git_dir <- dir.exists(file.path(current_dir, ".git"))

    if (has_source_csv || has_git_dir) {
      return(current_dir)
    }

    parent_dir <- dirname(current_dir)

    if (identical(parent_dir, current_dir)) {
      stop(
        "Could not find the project root. Run this report from the project folder or a subfolder.",
        call. = FALSE
      )
    }

    current_dir <- parent_dir
  }
}

project_root <- find_project_root()

csv_candidates <- c(
  file.path(project_root, "MSS_Master_Daily_S24_1980_2026.csv"),
  file.path(project_root, "data", "MSS_Master_Daily_S24_1980_2026.csv"),
  file.path(project_root, "data", "raw", "MSS_Master_Daily_S24_1980_2026.csv")
)

csv_path <- csv_candidates[file.exists(csv_candidates)][1]

if (is.na(csv_path)) {
  stop(
    "Could not find MSS_Master_Daily_S24_1980_2026.csv in the project root or data/raw/.",
    call. = FALSE
  )
}

rainfall_raw <- read_csv(
  csv_path,
  na = c("", "NA", "N/A", "-", "\u2014", "\u0097"),
  show_col_types = FALSE
)

rainfall_raw
```

------------------------------------------------------------------------

## 3 Clean the Rainfall Data

The raw file stores date information across separate year, month, and day columns. We clean the column names, convert the date fields to integers, create a proper `date` column, parse rainfall as a numeric variable, and trim whitespace from the station name.

```{r}
#| label: clean-rainfall-data
rainfall <- rainfall_raw |>
  clean_names() |>
  mutate(
    across(c(year, month, day), as.integer),
    date = make_date(year = year, month = month, day = day),
    daily_rainfall_total_mm = suppressWarnings(
      as.numeric(gsub("[^0-9.]", "", as.character(daily_rainfall_total_mm)))
    ),
    station = str_squish(station),
    month_name = factor(month.abb[month], levels = month.abb)
  ) |>
  filter(!is.na(daily_rainfall_total_mm)) |>
  arrange(date)

head(rainfall)
tail(rainfall)
```

### 3.1 Verify the Cleaned Data

These checks confirm the observed date range, duplicate dates, missing dates, missing rainfall values, and impossible negative rainfall values.

```{r}
#| label: verify-rainfall-cleaning
date_checks <- tibble(
  check = c(
    "rows",
    "first_date",
    "last_date",
    "invalid_dates",
    "duplicate_dates",
    "missing_dates_inside_observed_range",
    "missing_daily_rainfall_values",
    "negative_daily_rainfall_values"
  ),
  value = c(
    as.character(nrow(rainfall)),
    as.character(min(rainfall$date, na.rm = TRUE)),
    as.character(max(rainfall$date, na.rm = TRUE)),
    as.character(sum(is.na(rainfall$date))),
    as.character(sum(duplicated(rainfall$date))),
    as.character(length(setdiff(
      seq(
        min(rainfall$date, na.rm = TRUE),
        max(rainfall$date, na.rm = TRUE),
        by = "day"
      ),
      rainfall$date
    ))),
    as.character(sum(is.na(rainfall$daily_rainfall_total_mm))),
    as.character(sum(rainfall$daily_rainfall_total_mm < 0, na.rm = TRUE))
  )
)

date_checks
```

```{r}
#| label: verify-station-history
station_history <- rainfall |>
  group_by(station) |>
  summarise(
    first_date = min(date),
    last_date = max(date),
    n_days = n(),
    .groups = "drop"
  ) |>
  arrange(first_date)

station_change_dates <- rainfall |>
  mutate(previous_station = lag(station)) |>
  filter(!is.na(previous_station), station != previous_station) |>
  select(date, previous_station, station)

station_history
station_change_dates
```

The station history check confirms that these are station-level observations for Changi.

### 3.2 Export the Cleaned Daily Data

We write the cleaned daily dataset to `data/processed/` to ensure project reproducibility.

```{r}
#| label: export-cleaned-rainfall-data
dir.create(
  file.path(project_root, "data", "processed"),
  showWarnings = FALSE,
  recursive = TRUE
)

write_csv(
  rainfall,
  file.path(project_root, "data", "processed", "changi_daily_cleaned.csv")
)
```

------------------------------------------------------------------------

## 4 Derive Monthly Reference Data

We aggregate the daily rainfall observations into monthly totals as supporting reference data for the cleaning workflow. To handle incomplete data, we generate a complete year-month grid and classify each month as complete, partial, or unavailable. The final redesign in Section 5 uses daily rainfall directly because flood-stress thresholds are day-level events.

```{r}
#| label: derive-monthly-rainfall
monthly_observed <- rainfall |>
  mutate(month_start = floor_date(date, unit = "month")) |>
  group_by(year, month, month_start) |>
  summarise(
    monthly_total_mm = sum(daily_rainfall_total_mm, na.rm = TRUE),
    n_days_observed = n(),
    n_missing_rainfall = sum(is.na(daily_rainfall_total_mm)),
    expected_days = days_in_month(first(month_start)),
    .groups = "drop"
  ) |>
  mutate(
    month_status = case_when(
      n_days_observed == expected_days & n_missing_rainfall == 0 ~ "complete",
      TRUE ~ "partial"
    )
  )

all_months <- expand_grid(
  year = min(rainfall$year, na.rm = TRUE):max(rainfall$year, na.rm = TRUE),
  month = 1:12
) |>
  mutate(
    month_start = make_date(year, month, 1),
    expected_days = days_in_month(month_start)
  )

monthly <- all_months |>
  left_join(
    monthly_observed,
    by = c("year", "month", "month_start", "expected_days")
  ) |>
  mutate(
    month_status = case_when(
      is.na(n_days_observed) ~ "unavailable",
      month_status == "complete" ~ "complete",
      TRUE ~ "partial"
    ),
    month_name = factor(month.abb[month], levels = month.abb),
    year_display = factor(year, levels = rev(sort(unique(year)))),
    display_total_mm = if_else(
      month_status == "complete",
      monthly_total_mm,
      NA_real_
    )
  )

# Calculate historical baseline
climatology <- monthly |>
  filter(month_status == "complete") |>
  group_by(month_name) |>
  summarise(
    median_val = median(monthly_total_mm, na.rm = TRUE),
    q10 = quantile(monthly_total_mm, 0.10, na.rm = TRUE),
    q25 = quantile(monthly_total_mm, 0.25, na.rm = TRUE),
    q75 = quantile(monthly_total_mm, 0.75, na.rm = TRUE),
    q90 = quantile(monthly_total_mm, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

head(climatology)
```

------------------------------------------------------------------------

## 5 Redesigned Visualisations

This is the primary redesign chart for urban planners and flood-prevention stakeholders. Each column is one year and each row is one calendar month. Every cell in the grid covers one month in one year across the full 1983-2025 record.

The tile color encodes the number of heavy-rain days (daily rainfall ≥ 50 mm) in each month-year using a four-level discrete scale: 0 days (near-white), 1 day (light blue), 2 days (mid blue), 3 or more days (dark blue). Red circle markers appear over cells that contained at least one extreme-rain day (daily rainfall ≥ 100 mm); the size of the circle scales with the count of extreme days in that cell.

This keeps two planning signals visible at once: the seasonal clustering of heavy-rain months and isolated high-severity events that do not follow the seasonal pattern.

```{r}
#| label: prepare-year-month-flood-stress-matrix

daily_complete_43_years <- rainfall |>
  mutate(
    date = as.Date(date),
    year = as.integer(year),
    month = as.integer(month),
    daily_rainfall_total_mm = as.numeric(daily_rainfall_total_mm)
  ) |>
  filter(
    year >= 1983,
    year <= 2025
  )

daily_matrix_input_checks <- daily_complete_43_years |>
  summarise(
    first_date = min(date),
    last_date = max(date),
    n_years = n_distinct(year),
    n_days = n(),
    missing_month_values = sum(is.na(month)),
    missing_daily_rainfall_values = sum(is.na(daily_rainfall_total_mm))
  )

year_month_flood_stress <- daily_complete_43_years |>
  mutate(
    month_name = factor(month.abb[month], levels = rev(month.abb))
  ) |>
  group_by(year, month, month_name) |>
  summarise(
    heavy_days_50_99 = sum(
      daily_rainfall_total_mm >= 50 &
        daily_rainfall_total_mm < 100,
      na.rm = TRUE
    ),
    extreme_days_100_plus = sum(
      daily_rainfall_total_mm >= 100,
      na.rm = TRUE
    ),
    heavy_days_50_plus = heavy_days_50_99 + extreme_days_100_plus,
    max_daily_mm = max(daily_rainfall_total_mm, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    heavy_day_class = case_when(
      heavy_days_50_plus == 0 ~ "0",
      heavy_days_50_plus == 1 ~ "1",
      heavy_days_50_plus == 2 ~ "2",
      heavy_days_50_plus >= 3 ~ "3+",
      TRUE ~ NA_character_
    ),
    heavy_day_class = factor(
      heavy_day_class,
      levels = c("0", "1", "2", "3+")
    ),
    extreme_marker_size = if_else(
      extreme_days_100_plus > 0,
      as.numeric(extreme_days_100_plus),
      NA_real_
    )
  )

year_month_matrix_checks <- year_month_flood_stress |>
  summarise(
    n_years      = n_distinct(year),
    n_months     = n_distinct(month),
    n_cells      = n(),
    max_heavy_days_in_month_year   = max(heavy_days_50_plus),
    max_extreme_days_in_month_year = max(extreme_days_100_plus),
    missing_heavy_day_classes = sum(is.na(heavy_day_class))
  )

daily_matrix_input_checks
year_month_matrix_checks
year_month_flood_stress |>
  arrange(desc(heavy_days_50_plus), desc(extreme_days_100_plus), desc(max_daily_mm)) |>
  select(
    year,
    month_name,
    heavy_days_50_99,
    extreme_days_100_plus,
    heavy_days_50_plus,
    max_daily_mm
  ) |>
  head(15)
```

```{r}
#| label: plot-year-month-flood-stress-matrix
#| fig-width: 16
#| fig-height: 6

ggplot(
  year_month_flood_stress,
  aes(x = year, y = month_name)
) +
  geom_tile(
    aes(fill = heavy_day_class),
    color = "white",
    linewidth = 0.25
  ) +
  geom_point(
    data = year_month_flood_stress |>
      filter(extreme_days_100_plus > 0),
    aes(size = extreme_marker_size),
    shape = 21,
    fill = "#C2410C",
    color = "white",
    stroke = 0.45,
    alpha = 0.92
  ) +
  scale_fill_manual(
    values = c(
      "0"  = "#EEF2F7",
      "1"  = "#AECDE3",
      "2"  = "#4A9DC4",
      "3+" = "#0E4D82"
    ),
    drop = FALSE,
    name = "Heavy-rain days (\u226550 mm)",
    guide = guide_legend(
      direction = "horizontal",
      title.position = "top",
      label.position = "bottom",
      keywidth  = unit(1.4, "cm"),
      keyheight = unit(0.45, "cm")
    )
  ) +
  scale_size_continuous(
    range  = c(2.0, 5.8),
    breaks = c(1, 2, 3),
    limits = c(1, 3),
    name   = "Extreme days (\u2265100 mm)",
    guide  = guide_legend(
      direction    = "horizontal",
      title.position = "top",
      label.position = "bottom",
      override.aes = list(
        fill   = "#C2410C",
        color  = "white",
        shape  = 21,
        stroke = 0.45,
        alpha  = 0.92
      )
    )
  ) +
  scale_x_continuous(
    breaks = c(1983, seq(1985, 2025, by = 5)),
    expand = expansion(mult = c(0.005, 0.005))
  ) +
  labs(
    title    = "Year-Month Flood-Stress Matrix at Changi, 1983\u20132025",
    subtitle = "Each cell is one month in one year. Darker blue = more heavy-rain days (\u226550 mm); red circles = extreme-rain days (\u2265100 mm), sized by count.",
    x        = "Year",
    y        = NULL,
    caption  = paste0(
      "Data: Meteorological Service Singapore daily observations, Changi station.\n",
      "Heavy rain \u226550 mm/day; extreme rain \u2265100 mm/day. 2026 excluded (incomplete year)."
    )
  ) +
  guides(
    fill = guide_legend(
      order = 1,
      direction      = "horizontal",
      title.position = "top",
      label.position = "bottom",
      keywidth  = unit(1.4, "cm"),
      keyheight = unit(0.45, "cm")
    ),
    size = guide_legend(
      order = 2,
      direction      = "horizontal",
      title.position = "top",
      label.position = "bottom",
      override.aes = list(
        fill   = "#C2410C",
        color  = "white",
        shape  = 21,
        stroke = 0.45,
        alpha  = 0.92
      )
    )
  ) +
  theme(
    plot.title        = element_text(face = "bold", size = 16),
    plot.subtitle     = element_text(color = "grey25", size = 10.5),
    plot.caption      = element_text(color = "grey40", hjust = 0, size = 8.5),
    panel.grid        = element_blank(),
    axis.text.x       = element_text(size = 8.5, angle = 0, vjust = 0.5),
    axis.text.y       = element_text(size = 9.5, face = "bold"),
    axis.title.x      = element_text(face = "bold"),
    legend.position   = "bottom",
    legend.title      = element_text(face = "bold", size = 9),
    legend.text       = element_text(size = 8.5),
    legend.box        = "horizontal",
    legend.spacing.x  = unit(0.8, "cm"),
    plot.margin       = margin(10, 15, 10, 10)
  )
```

## Quick Reading

**Selected main redesign: Year-Month Flood-Stress Matrix.**

This chart was chosen because it preserves both planning dimensions at once: the full 43-year record and the month-by-month separation. Stakeholders can scan the poster and answer two questions at the same time — which months are seasonally risky, and what is the pattern for years that produced unusual events?


------------------------------------------------------------------------

## 7 Poster Notes

Key points to transfer to the poster:

1.  **Station-Level Specificity:** The data are station-level rainfall observations for Changi, not a national average for Singapore.
2.  **Date Range Coverage:** The observed date range is 1980-01-01 to 2026-05-31.
3.  **Complete-Year Comparison:** The matrix uses 1983-2025 only; 2026 is excluded because the source record stops at 2026-05-31.
4.  **Heavy-Rain Frequency:** Darker cells show months with repeated days at or above $50\text{ mm}$, making seasonal flood-stress clusters easier to scan.
5.  **Extreme-Rain Severity:** Red circles flag days at or above $100\text{ mm}$, so isolated high-severity events remain visible even when monthly heavy-rain frequency is low.

------------------------------------------------------------------------

## 8 References

1.  Chua Chin Hon (2019). "Visualising Singapore’s Changing Weather Patterns: 1983-2019". *Towards Data Science*. [Medium Article](https://medium.com/data-science/visualising-singapores-changing-weather-patterns-1983-2019-a78605fadbdf).
2.  Historical Daily Weather Records. Meteorological Service Singapore (MSS). [MSS Historical Data](http://www.weather.gov.sg/climate-historical-daily/).
