---
title: "Exploratory Data Analysis of Rainfall Data"
format: html
execute:
  warning: false
---

## Introduction

Exploratory data analysis of the rainfall data from the Changi station for Week 09 mid-project check in

```{r}
#| label: setup

library(dplyr)
library(ggplot2)
library(lubridate)
library(janitor)
library(tibble)
library(tidyr)
library(stringr)
library(viridis)
library(patchwork)
library(scales)
library(ggrepel)
library(zoo)

```

## Data Loading and Cleaning

The raw Changi station file covers 1980-2026. For the main comparison dataset, this
analysis uses 1983 onward. This matches the starting year of the reference Medium
visualisation, while still extending the time period beyond the original 1983-2019
range using the newer records available up to 2026.

The 1980-1982 records are not deleted from the raw source file. Instead, they are
cleaned first, checked for missingness, and then excluded from the main analysis
dataset. This makes the cleaning decision reproducible in R rather than relying on
manual spreadsheet edits.

```{r}
#| label: load-and-clean-data

# Create directories for processed data and figures
dir.create("data/processed", showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

# Define a function to parse numbers from character vectors
parse_base_number <- function(x) {
  cleaned <- gsub("[^0-9.\\-]+", "", as.character(x))
  cleaned[cleaned == ""] <- NA_character_
  suppressWarnings(as.numeric(cleaned))
}

# Find the CSV file
csv_path <- "MSS_Master_Daily_S24_1980_2026.csv"
if (!file.exists(csv_path)) {
  csv_path <- file.path("data", "raw", "MSS_Master_Daily_S24_1980_2026.csv")
}
if (!file.exists(csv_path)) {
  stop("Could not find MSS_Master_Daily_S24_1980_2026.csv")
}

# Read and clean the raw data
rainfall_raw <- read.csv(
  csv_path,
  na.strings = c("", "NA", "N/A", "-", "\u2014", "\u0097"),
  fileEncoding = "UTF-8",
  stringsAsFactors = FALSE,
  check.names = FALSE
)

rainfall_all <- rainfall_raw |>
  janitor::clean_names() |>
  mutate(
    across(c(year, month, day), as.integer),
    date = lubridate::make_date(year = year, month = month, day = day),
    across(
      c(
        daily_rainfall_total_mm,
        highest_30_min_rainfall_mm,
        highest_60_min_rainfall_mm,
        highest_120_min_rainfall_mm,
        mean_temperature_c,
        maximum_temperature_c,
        minimum_temperature_c,
        mean_wind_speed_km_h,
        max_wind_speed_km_h
      ),
      parse_base_number
    ),
    station = stringr::str_squish(station)
  ) |>
  arrange(date)

early_year_missingness <- rainfall_all |>
  filter(year >= 1980, year <= 1985) |>
  group_by(year) |>
  summarise(
    rows = n(),
    daily_rainfall_missing_percent = mean(is.na(daily_rainfall_total_mm)),
    highest_30_min_missing_percent = mean(is.na(highest_30_min_rainfall_mm)),
    highest_60_min_missing_percent = mean(is.na(highest_60_min_rainfall_mm)),
    highest_120_min_missing_percent = mean(is.na(highest_120_min_rainfall_mm)),
    temperature_fields_missing_percent = mean(
      is.na(mean_temperature_c) |
        is.na(maximum_temperature_c) |
        is.na(minimum_temperature_c)
    ),
    wind_fields_missing_percent = mean(
      is.na(mean_wind_speed_km_h) |
        is.na(max_wind_speed_km_h)
    ),
    .groups = "drop"
  )

print(early_year_missingness)

# Main comparison dataset:
# 1980-1982 are excluded because these years contain substantial missingness in
# supporting weather/intensity fields and the reference visualisation begins in
# 1983. Daily rainfall itself is complete in these years, so this is a design
# comparability decision rather than a claim that the rainfall totals are absent.
rainfall <- rainfall_all |>
  filter(year >= 1983, year <= 2026)

```

## Data Quality Checks

Here we perform some basic checks on the cleaned data to understand its integrity.

```{r}
#| label: data-quality-checks

date_checks <- tibble(
  check = c(
    "rows",
    "first_date",
    "last_date",
    "invalid_dates",
    "duplicate_dates",
    "missing_dates_inside_observed_range",
    "missing_daily_rainfall_values",
    "negative_daily_rainfall_values"
  ),
  value = c(
    as.character(nrow(rainfall)),
    as.character(min(rainfall$date, na.rm = TRUE)),
    as.character(max(rainfall$date, na.rm = TRUE)),
    as.character(sum(is.na(rainfall$date))),
    as.character(sum(duplicated(rainfall$date))),
    as.character(length(setdiff(
      seq(min(rainfall$date, na.rm = TRUE), max(rainfall$date, na.rm = TRUE), by = "day"),
      rainfall$date
    ))),
    as.character(sum(is.na(rainfall$daily_rainfall_total_mm))),
    as.character(sum(rainfall$daily_rainfall_total_mm < 0, na.rm = TRUE))
  )
)

print(date_checks)

field_missingness <- rainfall |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "column",
    values_to = "missing_values"
  ) |>
  mutate(missing_percent = missing_values / nrow(rainfall))

print(field_missingness)
```

## Data Aggregation

We will aggregate the daily data into monthly and yearly summaries. These summaries are useful for understanding the data at different temporal granularities and for later analysis.

```{r}
#| label: aggregate-data

monthly_observed <- rainfall |>
  mutate(month_start = floor_date(date, unit = "month")) |>
  group_by(year, month, month_start) |>
  summarise(
    monthly_total_mm = sum(daily_rainfall_total_mm, na.rm = TRUE),
    n_days_observed = n(),
    n_missing_rainfall = sum(is.na(daily_rainfall_total_mm)),
    expected_days = lubridate::days_in_month(first(month_start)),
    .groups = "drop"
  ) |>
  mutate(
    month_status = case_when(
      n_days_observed == expected_days & n_missing_rainfall == 0 ~ "complete",
      TRUE ~ "partial"
    )
  )

all_months <- tidyr::expand_grid(
  year = min(rainfall$year, na.rm = TRUE):max(rainfall$year, na.rm = TRUE),
  month = 1:12
) |>
  mutate(
    month_start = lubridate::make_date(year, month, 1),
    expected_days = lubridate::days_in_month(month_start)
  )

monthly <- all_months |>
  left_join(
    monthly_observed,
    by = c("year", "month", "month_start", "expected_days")
  ) |>
  mutate(
    month_status = case_when(
      is.na(n_days_observed) ~ "unavailable",
      month_status == "complete" ~ "complete",
      TRUE ~ "partial"
    ),
    month_name = factor(month.abb[month], levels = month.abb),
    year_display = factor(year, levels = rev(sort(unique(year)))),
    display_total_mm = if_else(month_status == "complete", monthly_total_mm, NA_real_)
  )

annual <- monthly |>
  group_by(year) |>
  summarise(
    observed_total_mm = sum(monthly_total_mm, na.rm = TRUE),
    complete_months = sum(month_status == "complete"),
    unavailable_months = sum(month_status == "unavailable"),
    partial_months = sum(month_status == "partial"),
    annual_total_mm = if_else(
      complete_months == 12,
      observed_total_mm,
      NA_real_
    ),
    year_status = if_else(complete_months == 12, "complete", "partial_or_unavailable"),
    .groups = "drop"
  )

# Displaying the first few rows of the aggregated data
print("Monthly Data Head:")
head(monthly)

print("Annual Data Head:")
head(annual)
```

## Save Processed Data

The cleaned and aggregated data are saved to CSV files. This fulfills the requirement to provide the dataset in its cleaned form.

```{r}
#| label: save-processed-data

write.csv(rainfall_all, file.path("data", "processed", "changi_daily_cleaned_full_1980_2026.csv"), row.names = FALSE, na = "")
write.csv(rainfall, file.path("data", "processed", "changi_daily_cleaned.csv"), row.names = FALSE, na = "")
write.csv(monthly, file.path("data", "processed", "changi_monthly_rainfall.csv"), row.names = FALSE, na = "")
write.csv(annual, file.path("data", "processed", "changi_annual_rainfall.csv"), row.names = FALSE, na = "")

```

---

## Design Brief Draft

(TODO: Add your 50-125 word design brief here.)

---

## Planned Design Direction

(TODO: Add your 1-2 paragraph note on your planned design direction and any questions for the instructors here.)

---
